Supplement for: Carbon cycling in mature and regrowth forests globally: a macroecological synthesis based on the global Forest Carbon (ForC) database

### 2021-01-21 manuscript/ERL_global_C_review.Rmd
worked on intro

#figures
schematic =1
figure_map=2
diagram_TrB_mature= 3
diagram_TeB_mature=4
diagram_TeN_mature=5
diagram_BoN_mature=6
all_diagrams_mature="3-6"
flux_age_trends=7
stock_age_trends=8
TrB_age_trends=9
TeB_age_trends=10
TeN_age_trends=11
BoN_age_trends=12

@@

For more than half a century, researchers have sought to understand how forest carbon cycling varies across stands, including those of different biomes [e.g., @lieth_primary_1973; @luyssaert_co2_2007] and stand ages [e.g., @odum_strategy_1969; @luyssaert_old-growth_2008].
Over this time, an increasingly refined conceptual understanding of the elements of ecosystem C cycles has developed, as a growing number of variables have been defined [e.g., @chapin_reconciling_2006], along with appropriate measurement methods [e.g., @clark_measuring_2001]. 
New technology has also enabled researches to directly measure an expanding set of variables, notably including the development of continuous measurements of soil CO~2~ efflux (**REF-BEN**) and ecosystem-atmosphere CO~2~ exchange (**REFS**).
Standardization of estimation techniques has also increased; for example, there has been increasing standardization of the biomass allometries strongly influence estimates of any C cycle variable including woody biomass or changes therein [e.g., @chojnacky_updated_2014; @chave_improved_2014].
Further standardization has been made possible through research networks such as ForestGEO [@anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021], NEON [@schimel_neon_2007], and FLUXNET [@baldocchi_fluxnet_2001; @novick_ameriflux_2018].
Remote sensing technology has become increasingly useful for global- or regional-scale estimates of a few critical variables [e.g., aboveground biomass: @hu_mapping_2016; @spawn_harmonized_2020; gross primary productivity, $GPP$: @li_mapping_2019; @saatchi_benchmark_2011], yet measurement and validation of most forest C stocks and fluxes necessarily requires intensive on-the-ground data collection.

@@

Forest C cycling varies enormously across biomes, which are artificial categorical distinctions convenient for categorizing the world's forests according to major differences in climate, vegetation, *etc.* 
Since the early 19th century, it has been recognized that climate plays a dominant role in shaping differences among forests on a global scale [@humboldt_essay_1807; @holdridge_determination_1947].
*C stocks*
Early quantitative syntheses revealed that C fluxes such as net primary productivity ($NPP$) decrease dramatically with latitude, or increase across gradients of mean annual temperature and, to a lesser extent, precipitation [@lieth_primary_1973].

@banbury_morgan_global_nodate

Recently, machine learning draws on global maps in combination with ground-based data to create global maps of C cycling [e.g., REFS; @cook-patton_mapping_2020].

While current data and technology allow increasingly fine-scale resolution for understanding differences across forests, categorization into biomes remains useful for applications where data do not yet support robust analysis across finer-scale axes of variation.

### Age trends and their variation across biomes
In 1969, Eugene Odum published ... [@odum_strategy_1969].
Although the conceptualization of the C cycle in this paper is simplistic by current standards, the paper was foundational... [@corman_foundations_2019].
...
*(Need for a synthetic update.)*

In the past *couple of/several* decades, researchers have started asking how age trends--mostly in biomass accumulation-- vary across biomes.
*(look up early work on this--citations in @anderson_temperature-dependence_2006)*
@anderson_temperature-dependence_2006
@chazdon_carbon_2016
@cook-patton_mapping_2020
There has been little synthesis of differences in variables other than C stocks or their accumulation rates.
Given the important role of secondary forests in the current and future global C cycle, concrete understanding of age trends in C fluxes and stocks and how these vary across biomes is critical to better understanding of the global C cycle.

```{r schematic, eval = generate_figures, echo=FALSE, fig.align = "center", fig.cap = "Figure 1 | Schematic diagram summarizing current understanding of biome differences and age trends in forest C cycling. Variables are defined in Table 1. Age trends, which represent idealized dynamics following a disturbance that removes all living and non-living vegetation, are an updated version of the classic figure from Odum (1969), with heavy lines corresponding to those in Odum's figure and $NEP$ corresponding to Odum's 'net production'. Here, $NEP$ conists primarily of woody aboveground net primary production ($ANPP_{woody}$), while $ANPP_{woody.turnover}$ is the sum of woody mortality and branch turnover ", out.width='100%', fig.pos='H'}

>>>>

#figures
schematic =1
schematic_fluxes_biome="1a"
  schematic_fluxes_age="1b"
  schematic_stocks_biome="1c"
  schematic_stocks_age="1d"
  schematic_fs_age = "1b,d"
figure_map=2
diagram_TrB_mature= 3
diagram_TeB_mature=4
diagram_TeN_mature=5
diagram_BoN_mature=6
all_diagrams_mature="3-6"
flux_age_trends=7
stock_age_trends=8
TrB_age_trends=9
TeB_age_trends=10
TeN_age_trends=11
BoN_age_trends=12

@@

For more than half a century, researchers have sought to understand how forest carbon cycling varies across stands, including those of different biomes [e.g., @lieth_primary_1973; @luyssaert_co2_2007] and stand ages [e.g., @odum_strategy_1969; @luyssaert_old-growth_2008].
Over this time, an increasingly refined conceptual understanding of the elements of ecosystem C cycles has developed, as a growing number of variables have been defined [e.g., @chapin_reconciling_2006], along with appropriate measurement methods [e.g., @clark_measuring_2001]. 
New technology has also enabled researches to directly measure an expanding set of variables, notably including the development of continuous measurements of soil CO~2~ efflux (**REF-BEN**) and ecosystem-atmosphere CO~2~ exchange (**REFS**).
Standardization of estimation techniques has also increased; for example, there has been increasing standardization of the biomass allometries strongly influence estimates of any C cycle variable including woody biomass or changes therein [e.g., @chojnacky_updated_2014; @chave_improved_2014].
Further standardization has been made possible through research networks such as ForestGEO [@anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021], NEON [@schimel_neon_2007], and FLUXNET [@baldocchi_fluxnet_2001; @novick_ameriflux_2018].
Remote sensing technology has become increasingly useful for global- or regional-scale estimates of a few critical variables [e.g., aboveground biomass, $B_{ag}$: @saatchi_benchmark_2011; @hu_mapping_2016; @spawn_harmonized_2020; gross primary productivity, $GPP$: @li_mapping_2019], yet measurement and validation of most forest C stocks and fluxes necessarily requires intensive on-the-ground data collection.

@@

Forest C cycling varies enormously across biomes, which are artificial categorical distinctions convenient for categorizing the world's forests according to major differences in climate, vegetation, *etc.* 
Since the early 19th century, it has been recognized that climate plays a dominant role in shaping differences among forests on a global scale [@humboldt_essay_1807; @holdridge_determination_1947].
Global scale data synthesis has revealed that C fluxes including $GPP$, net primary productivity ($NPP$), and soil respiration ($R_{soil}$) decrease with latitude or, correspondingly, increase with mean annual temperature and, to a lesser extent, precipitation [Fig. `r schematic_fluxes_biome` **REFS**; @lieth_primary_1973; @luyssaert_co2_2007; @banbury_morgan_global_nodate].
C stocks of mature forests show less directional variation (Fig. `r schematic_fluxes_biome`).
On average, $B_{ag}$ tends to decrease with latitude, but not as dramatically as fluxes, and with the highest $B_{ag}$ forests in relatively cool, moist temperate regions [**REFS**; @keith_re-evaluation_2009]. 
In contrast, standing and downed dead wood ($DW_{standing}$ and $DW_{down}$, respectively) and the organic layer ($OL$) tend to accumulate more in colder climates where decomposition is slow relative to $NPP$ [**REFS**].

Correlative analyses relating C cycle variables to climate and other environmental variables have recently been taken to a new level through use of machine-learning algorithms that relate ground-based C cycle data to *global data on climate/soils/satellite data*, making it possible to create fine-scale global maps of C cycling [e.g., **REFS**; @cook-patton_mapping_2020].
*This approach can be particularly effective when paired with satellite data ...  (e.g., aboveground biomass: Saatchi et al 2011, Hu et al 2016, Spawn et al 2020, gross primary productivity, GPP: Li and Xiao 2019, ).* 
Yet, any such analysis is constrained by the quality and coverage of ground-based estimates of forest C fluxes or stocks. 
While estimates of some variables (e.g., $B_{ag}$, $GPP$, $NPP$, $R_{soil}$) are widely available, many remain poorly characterized (e.g., $DW$; autotrophic respiration, $R_{auto}$) --even at the coarse resolution of biomes.
This is a critical limitation not only for understanding forest C cycling, but also for quantifying climate change mitigation across forest biomes or ecozones [e.g., **REFS**].

### Age trends and their variation across biomes

Stand age is another important axis of variation in forest C cycling (Fig. `r schematic_fs_age`).
In 1969, E.P. Odum's "The Strategy of Ecosystem Development" laid out predictions as to how forest energy flows and organic matter stocks vary with stand age  [@odum_strategy_1969].
Although the conceptualization of the C cycle in this paper is simplistic by current standards, the paper was foundational in framing the theory around which research on the subject still revolves [@corman_foundations_2019], and the basic framework still holds, *albiet* with modest modifications [@andersonteixeira_altered_2013].
Following stand-clearing disturbance, $GPP$, $NPP$, and biomass of leaves ($B_{foliage}$) and fine roots ($B_{root-fine}$) increase rapidly and thereafter remain relatively stable ($B_{foliage}$,$B_{root-fine}$, sometimes $GPP$) or decline slightly [$NPP$, sometimes $GPP$; e.g., *Goulden et al. 2011, ....  refs in * @andersonteixeira_altered_2013].
The decline in $NPP$ occurs because $R_{auto}$ increases relative to $GPP$ as forests age, corresponding to declining carbon use efficiency with stand age [@delucia_forest_2007; @collalti_forest_2020].
Heterotrophic respiration, most of which originates from the soil ($R_{het-soil}$) remains relatively constant with stand age *[Law et al., 2003; Pregitzer & Euskirchen, 2004; Goulden et al., 2011]*, with the result that net ecosystem production ($NEP=GPP-R_{eco}$, where $R_{eco}$ is total ecosystem respiration) is initially negative, increases to a maximum at intermediate ages, and declines thereafter *[Law et al., 2003; Pregitzer & Euskirchen, 2004; Zhou et al., 2006; Baldocchi, 2008; Luyssaert et al., 2008; Amiro et al., 2010; Goulden et al., 2011]*.
The result is that biomass accumulates rapidly in young forests, followed by a slow decline to near zero in old forests *[Lichstein et al., 2009; Yang et al., 2011;
Hember et al., 2012]*.
While these trends have been subject of fairly recent review [@andersonteixeira_altered_2013], *there is need for a synthetic, quantitaitive review.*

In the past *couple of/several* decades, researchers have started asking how age trends--mostly in biomass accumulation-- vary across biomes.
*(look up early work on this--citations in @anderson_temperature-dependence_2006)*
@anderson_temperature-dependence_2006
@chazdon_carbon_2016, 
Tepley et al., 
@cook-patton_mapping_2020
There has been little synthesis of differences in variables other than C stocks or their accumulation rates.
Given the important role of secondary forests in the current and future global C cycle, concrete understanding of age trends in C fluxes and stocks and how these vary across biomes is critical to better understanding of the global C cycle.

```{r schematic, eval = generate_figures, echo=FALSE, fig.align = "center", fig.cap = "Figure 1 | Schematic diagram summarizing current understanding of biome differences and age trends in forest C cycling. Variables are defined in Table 1. Age trends, which represent idealized dynamics following a disturbance that removes all living and non-living vegetation, are an updated version of the classic figure from Odum (1969), with heavy lines corresponding to those in Odum's figure and $NEP$ corresponding to Odum's 'net production'. Here, $NEP$ conists primarily of woody aboveground net primary production ($ANPP_{woody}$), while $ANPP_{woody.turnover}$ is the sum of woody mortality and branch turnover. ", out.width='100%', fig.pos='H'}
